LME-MPM applied to quasi-brittle fracture.

### 2020-04-22 _region_.tex
Versión 22-4-2020

\documentclass[preprint,12pt,a4paper]{elsarticle}

@@

\usepackage{amsmath,amsthm,amssymb}

@@

@@

\begin{abstract}

  The objective of this work is to introduce an alternative
  technique to address the fracture process of brittle and
  quasi-brittle materials under the material point method (MPM)
  framework. With this purpose the eigensoftening algorithm, developed
  originally for the optimal transportation meshfree (OTM)
  approximation scheme, is extended to the MPM with the aim of present
  a suitable alternative to the existing fracture algorithms developed
  for the MPM. The good fitting in the predictions made by the
  eigensoftening algorithm against both analytical and experimental
  results proofs the well performance of the method under challenging loads.

\end{abstract}

@@

\subsection{Spatial discretization : Local-\textit{max-ent} approximants}
\label{sec:2.2}

Local maximum-entropy (or local \textit{max-ent}) approximation scheme
was introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} as a bridge between finite elements and meshfree methods. The key idea of the  shape functions is to interpret the nodal of a shape function $N_I$ as
a probability. This allow us to introduce two important limits:
the principle of maximum-entropy (\textit{max-ent}) statistical
inference stated by \cite{Jaynes1957}, and the Delaunay triangulation
which ensures the minimal width of the shape function. To reach to a compromise between two competing objectives, a Pareto set is defined as, 
\begin{align*}
  \label{eq:LME-scheme-pareto-set}
  \text{(LME)}_{\beta} \hspace{0.15cm} &\text{For fixed} \hspace{0.15cm}
  \vec{x} \hspace{0.15cm} \text{minimise} \hspace{0.15cm} f_{\beta}(\vec{x}_p, N_I) = \beta U(\vec{x}_p,N_I) - H(N_I) \\
  &\text{subject to}\
  \begin{cases}
    N_I \ge 0, \hspace{0.15cm} \text{I=1, ..., n} \\[1em]   
    \sum\limits_{I=1}^{N_n}{N_I} = 1 \\[1em]   
    \sum\limits_{I=1}^{N_n}{N_I \vec{x}_I} = \vec{x} \\
  \end{cases}
\end{align*}
where $H(N_I)$ is the entropy of the system of nodes following the
definition given by Shannon (1948) \cite{Shannon1948}, and $U(\vec{x}_p,N_I) =
\sum_I N_I |\vec{x}_p - \vec{x}_I |^2$ a magnitude of the shape
function width. The regularization o \textit{thermalization} parameter
between the two criterion $\beta$ has Pareto optimal values in the range
$(0,\infty)$. The unique solution of the local max-ent problem
(LME)$_\beta$ is:
\begin{equation}
  \label{eq:LME-p}
N_I^*(\vec{x})=\frac{\exp\left[ -\beta \; |\vec{x}-\vec{x}_I|^2 +
    \vec{\lambda}^* \cdot (\vec{x}-\vec{x}_I) \right] } {Z(\vec{x},\vec{\lambda}^*(\vec{x}))}
\end{equation}
where $Z(\vec{x},\vec{\lambda}^*(\vec{x}))$ is the \textit{partition
  function} defined as,
\begin{equation}
  \label{eq:LME-Z}
Z(\vec{x}, {\vec{\lambda}}) = \sum_{I=1}^{N_n}{ \exp \left[ -\beta \; |\vec{x}-\vec{x}_I|^2 + \vec{\lambda} \cdot (\vec{x}-\vec{x}_I)  \right]}
\end{equation}
and evaluated in the unique minimiser $\vec{\lambda}^*(\vec{x})$ for
the function $\log
Z(\vec{x}, \vec{\lambda})$. The traditional way to obtain such a
minimiser is using (\ref{eq:LME-J}) to calculate small increments of $\partial\vec{\lambda}$ in a
Newton-Raphson approach. Where $\tens{J}$ is the Hessian matrix, defined by:
\begin{eqnarray}
  \label{eq:LME-J} 
  \tens{J}(\vec{x}, \vec{\lambda},\beta) &\equiv& \frac{\partial
                                                  \vec{r}}{\partial \vec{\lambda}}\\
  \label{eq:LME-r}
  \vec{r}(\vec{x},\vec{\lambda},\beta) &\equiv& \frac{\partial \log{ Z(   \vec{x},\vec{\lambda}})}{\partial \vec{\lambda}}  = \sum_I^{N_n} p_I(\vec{x},\vec{\lambda},\beta) \, (\vec{x} - \vec{x}_I)
\end{eqnarray}
In order to obtain the first derivatives of the shape function~$\nabla
N^*_I$, can be computed as,
\begin{equation}
  \label{eq:LME-grad-p}
\nabla N^*_I = N^*_I  \, \left(\nabla f^*_I-\sum_J^{N_n} N^*_J \, \nabla f^*_J\right)
\end{equation}
where
\begin{equation}
  \label{eq:LME-f}
f^*_I(\vec{x},  \vec{\lambda},\beta)=-\beta \, |\vec{x}-\vec{x}_I|^2 + \vec{\lambda}^*  \,  (\vec{x}-\vec{x}_I)
\end{equation}
Employing the chain rule over \eqref{eq:LME-grad-p}, rearranging and considering $\beta$ as a
constant, Arroyo and Ortiz~\cite{Arroyo2006} obtained the following
expression for the gradient of the shape function.
\begin{eqnarray}
  \label{eq:LME-gradp} 
\nabla N^*_I &=& -N^*_I \,  (\tens{J}^*)^{-1} \,  (\vec{x} - \vec{x}_I)
\end{eqnarray}

The regularization parameter $\beta$ of LME shape functions may be
controlled by adjusting a dimensionless parameter, $\gamma=\beta h^2$
\cite{Arroyo2006}, where $h$ is defined as a measure of the nodal
spacing. Since  $N_I$ is defined in the entire domain, in practice,
the shape function decay $\exp(-\beta \vec{r} )$ is truncated  by  a
given tolerance, 10$^{-6}$, for example,  would ensure a reasonable range of
neighbours, see \cite{Arroyo2006} for details. This tolerance defines
the limit values of the influence radius and is used thereafter to
find the neighbour nodes of a given integration point.

\subsection{Fracture modelling approach}
\label{sec:2.3}

Within the context of MPM formulation, fracture can be modelled by
failing particles according to a suitable criterion. When material
points are failed, they are assumed to have a null stress
tensor. Navas {\it et al.} (2017)\cite{Navas_2017_ES} developed a
eigensoftening algorithm as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. A
comparison between both in \cite{Navas_2017_ES} shows that
the eigenerosion algorithm significantly overestimates the tensile
stress and the strain peaks, while it captures the forces and crack
patterns accurately. On the other hand eigensoftening algorithm agree
very well with experimental results in all the aspects. Furthermore,
this algorithm has also proof its accuracy for complex fracture
patters such the present in fiber reinforces concrete (FRC),
\cite{Navas_2018_ES}.\\

@@

As can be noticed, in the eigenerosion algorithm an energetic
criterion is adopted. Due to that fact, unrealistic stress
concentration (higher than tensile strength) in quasi-brittle materials, see
\cite{Navas_2017_ES}. To overcome this limitation, the aforementioned
authors proposed the concept of eigensoftening to take in to account
the gradual failure in quasi-brittle materials. The concept is
inspired in the cohesive fracture widely employed in the context of
FEM \cite{Ortiz_1999}. This gradual failure criterion is plotted in
figure \label{fig:Damage-ft-wc}, where a linear decreasing cohesive
law is presented to illustrate the concept here described. In the
picture, the shaded area represents the static fracture energy per
unit of area, $G_F$. As we can see, a cohesive crack appears when the
maximum tensile strength, $f_t$ is reached. Once the opening
displacement $w$ takes the value of the critical crack displacement
$w_c$, a stress-free crack is attained. For intermediate values,
$w_n$, a damage value between zero and one represents the extension to
which the material has failed.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/Damage}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Damage-ft-wc}
\end{figure}
%%%%%%%%%%%
For the eigensoftening algorithm,

a strength criterion for crack initialization was adopted

\begin{equation}
  \label{eq:variation-averaged-strain-energy-density}
  \delta W_{p,\epsilon} = \frac{\partial G_p}{C_{\epsilon}} =
  \frac{1}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} \delta \tens{\epsilon}_q
\end{equation}

\begin{equation}
  \label{eq:variation-averaged-strain-energy-density-simpli}
  \delta W_{p,\epsilon} =
  \frac{\delta \tens{\epsilon}_p}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} 
\end{equation}

\begin{equation}
  \label{eq:equivalent-critical-stress}
  \delta \tens{\sigma}_{p,c} =
  \frac{1}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} 
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cases of study and discussion}
\label{sec:3}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/Brazilian}
  \caption{Geometry and boundary condition of the Brazilian test.}
  \label{fig:geometry-brazilian-test}
\end{figure}

@@

\begin{algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit Predictor-Corrector scheme}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Update mass matrix}:
    \begin{equation*}
      \tens{m}_{I} = N_{Ip}^{k}\ m_p,
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Predictor}:\\
    \begin{equation*}
      \vec{v}_I^{pred} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{m_I}
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions}:\\
    At the fixed boundary, set $\vec{v}_{I}^{pred} = 0$. 
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    % \STATE \textbf{Discard the previous nodal values}.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{align*}
      &\dot{\tens{\varepsilon}_{p}}^{k+1} = \left[ \vec{v}_{I}^{pred} \otimes
        \Grad{N_{Ip}^{k+1}} \right]^s \\
      &\Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\ \dot{\tens{\varepsilon}_{p}}^{k+1}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Compute damage parameter}:
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    (1-\chi)\vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$ evaluating
    \eqref{eq:nodal_internal_forces} and
    \eqref{eq:nodal_external_forces} in the time step $k+1$.
    In the grid node $I$ is fixed in one of the spatial dimensions, set it to
    zero to make the grid nodal acceleration zero in that direction.\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector}:
    \begin{equation*}
      \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\ \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \STATE \textbf{Update particles lagrangian quantities}:
    \begin{align*}
      &\vec{a}_p^{k+1} = \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Reset nodal values}
  \end{algorithmic}
\end{algorithm}

@@

\end{document}

>>>>

\message{ !name(2020_EFM_MPM_Eigensoftening.tex)}\documentclass[preprint,12pt,a4paper]{elsarticle}

@@

\usepackage{amsmath,amsthm,amssymb}
\usepackage{xstring}

@@

\message{ !name(2020_EFM_MPM_Eigensoftening.tex) !offset(-3) }

@@

\begin{abstract}
  The objective of this work is to introduce an alternative
  technique to address the fracture process of brittle and
  quasi-brittle materials under the material point method (MPM)
  framework. With this purpose the eigensoftening algorithm, developed
  originally for the optimal transportation meshfree (OTM)
  approximation scheme, is extended to the MPM with the aim of present
  a suitable alternative to the existing fracture algorithms developed
  for the MPM. The good fitting in the predictions made by the
  eigensoftening algorithm against both analytical and experimental
  results proofs the well performance of the method under challenging loads.
\end{abstract}

@@

\subsection{Spatial discretization : Local-\textit{max-ent} approximants}
\label{sec:2.2}
Local maximum-entropy (or local \textit{max-ent}) approximation scheme
was introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} as a bridge
between finite elements and meshfree methods. The key idea of the
shape functions is to interpret the nodal of a shape function $N_I$ as
a probability. This allow us to introduce two important limits:
maximum-entropy (\textit{max-ent}) limit, and the Delaunay triangulation
which ensures the minimal width of the shape function. The first
limit ensures a \textit{unbiased statistical inference} based on the
nodal data as ensures the Jayne's\cite{Jaynes1957} principle of
\textit{maximum entropy}, the second limit warrant the \textit{least
  width} shape function support. To reach to a compromise between two
competing objectives, a Pareto set is defined as, 
\begin{align*}
  \label{eq:LME-scheme-pareto-set}
  \text{(LME)}_{\beta} \hspace{0.15cm} &\text{For fixed} \hspace{0.15cm}
  \vec{x} \hspace{0.15cm} \text{minimise} \hspace{0.15cm} f_{\beta}(\vec{x}_p, N_I) = \beta U(\vec{x}_p,N_I) - H(N_I) \\
  &\text{subject to}\
  \begin{cases}
    N_I \ge 0, \hspace{0.15cm} \text{I=1, ..., n} \\[1em]   
    \sum\limits_{I=1}^{N_n}{N_I} = 1 \\[1em]   
    \sum\limits_{I=1}^{N_n}{N_I \vec{x}_I} = \vec{x} \\
  \end{cases}
\end{align*}
where $H(N_I)$ is the entropy of the system of nodes following the
definition given by Shannon (1948) \cite{Shannon1948}, and $U(\vec{x}_p,N_I) =
\sum_I N_I |\vec{x}_p - \vec{x}_I |^2$ a 
magnitude of the shape function width. The regularization o
\textit{thermalization} parameter between the two criterion $\beta$
has Pareto optimal values in the range $(0,\infty)$. The unique
solution of the local max-ent problem (LME)$_\beta$ is
 \begin{equation}
  \label{eq:LME-p}
N_I^*(\vec{x})=\frac{\exp\left[ -\beta \; |\vec{x}-\vec{x}_I|^2 +
    \vec{\lambda}^* \cdot (\vec{x}-\vec{x}_I) \right] } {Z(\vec{x},\vec{\lambda}^*)}
\end{equation}
where $Z(\vec{x},\vec{\lambda}^*)$ is the \textit{partition function} defined as,
\begin{equation}
  \label{eq:LME-Z}
Z(\vec{x}, {\vec{\lambda}}) = \sum_{I=1}^{N_n}{ \exp \left[ -\beta \; |\vec{x}-\vec{x}_I|^2 + \vec{\lambda} \cdot (\vec{x}-\vec{x}_I)  \right]}
\end{equation}
and evaluated in the unique minimiser $\vec{\lambda}^*$ for the function $\log
Z(\vec{x}, \vec{\lambda})$. The traditional way to obtain
such a minimiser is using (\ref{eq:LME-J}) to calculate small
increments of $\partial\vec{\lambda}$ in a Newton-Raphson
approach. Where $\tens{J}$ is the Hessian matrix, defined by:
\begin{eqnarray}
  \label{eq:LME-J} 
  \tens{J}(\vec{x}, \vec{\lambda},\beta) &\equiv& \frac{\partial
                                                  \vec{r}}{\partial \vec{\lambda}}\\
  \label{eq:LME-r}
  \vec{r}(\vec{x},\vec{\lambda},\beta) &\equiv& \frac{\partial \log{ Z(   \vec{x},\vec{\lambda}})}{\partial \vec{\lambda}}  = \sum_I^{N_n} p_I(\vec{x},\vec{\lambda},\beta) \, (\vec{x} - \vec{x}_I)
\end{eqnarray}
The value of $\tens{J}$ can be employed to get the first derivatives of the shape function $\nabla
N^*_I$,
\begin{equation}
  \label{eq:LME-grad-p}
\nabla N^*_I = N^*_I  \, \left(\nabla f^*_I-\sum_J^{N_n} N^*_J \, \nabla f^*_J\right)
\end{equation}
where
\begin{equation}
  \label{eq:LME-f}
f^*_I(\vec{x},  \vec{\lambda},\beta)=-\beta \, |\vec{x}-\vec{x}_I|^2 + \vec{\lambda}^*  \,  (\vec{x}-\vec{x}_I)
\end{equation}
Employing the chain rule over \eqref{eq:LME-grad-p}, rearranging and considering $\beta$ as a
constant, Arroyo and Ortiz~\cite{Arroyo2006} obtained the following
expression for the gradient of the shape function.
\begin{eqnarray}
  \label{eq:LME-gradp} 
\nabla N^*_I &=& -N^*_I \,  (\tens{J}^*)^{-1} \,  (\vec{x} - \vec{x}_I)
\end{eqnarray}
The regularization parameter $\beta$ of LME shape functions may be
controlled by adjusting a dimensionless parameter, $\gamma=\beta h^2$
\cite{Arroyo2006}, where $h$ is defined as a measure of the nodal
spacing. Since  $N_I$ is defined in the entire domain, in practice,
the shape function decay $\exp(-\beta \vec{r} )$ is truncated  by  a
given tolerance, 10$^{-6}$, for example,  would ensure a reasonable
range of neighbours, see \cite{Arroyo2006} for details. This tolerance defines
the limit values of the influence radius and is used thereafter to
find the neighbour nodes of a given integration point.

\subsection{Fracture modelling approach}
\label{sec:2.3}
Within the context of MPM formulation, fracture can be modelled by
failing particles according to a suitable criterion. When material
points are failed, they are assumed to have a null stress
tensor. Navas {\it et al.} (2017)\cite{Navas_2017_ES} developed a
eigensoftening algorithm as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. A
comparison between both in \cite{Navas_2017_ES} shows that
the eigenerosion algorithm significantly overestimates the tensile
stress and the strain peaks, while it captures the forces and crack
patterns accurately. On the other hand eigensoftening algorithm agree
very well with experimental results in all the aspects. Furthermore,
this algorithm has also proof its accuracy for complex fracture
patters such the present in fiber reinforces concrete (FRC),
\cite{Navas_2018_ES}.\\

@@

As can be noticed, in the eigenerosion algorithm an energetic
criterion is adopted. Due to that fact, unrealistic stress
concentration (higher than tensile strength) in quasi-brittle materials, see
\cite{Navas_2017_ES}. To overcome this limitation, the aforementioned
authors proposed the concept of eigensoftening to take in to account
the gradual failure in quasi-brittle materials. The concept is
inspired in the cohesive fracture widely employed in the context of
FEM \cite{Ortiz_1999}. This gradual failure criterion is plotted in
figure \label{fig:Damage-ft-wc}, where a linear decreasing cohesive
law is presented to illustrate the concept here described. In the
picture, the shaded area represents the static fracture energy per
unit of area, $G_F$. As we can see, a cohesive crack appears when the
maximum tensile strength, $f_t$ is reached. Once the opening
displacement $w$ takes the value of the critical crack displacement
$w_c$, a stress-free crack is attained. For intermediate values,
$w_n$, a damage value between zero and one represents the extension to
which the material has failed.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/Damage}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Damage-ft-wc}
\end{figure}
%%%%%%%%%%%
For the eigensoftening algorithm, a strength criterion for crack
initialization was adopted. Particularly the maximum principal stress
theory for brittle fracture was considered by authors
\cite{Navas_2017_ES}. With that purpose, the variation of the averaged strain
energy density in the $\epsilon$-neighborhood of the material point
$\vec{x}_p^{k+1}$ can be obtained as,
\begin{equation}
  \label{eq:variation-averaged-strain-energy-density}
  \delta W_{\epsilon,p} = \frac{\partial G_p}{C_{\epsilon}} =
  \frac{1}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} \delta \tens{\epsilon}_q
\end{equation}
where $\tens{\sigma}_{q,I}$ is the maximum principal stress of each
material point in the $\epsilon$-neighborhood. Here
\cite{Navas_2017_ES} introduces an effective strain $\varepsilon_q$,
such the variation of the local strain energy can be obtained as
$\delta W_q = \sigma_{q,1} \delta\varepsilon_q$. Now with the assumption
that the effective strain of each material point at every time step
is constant in the neighborhood of $\vec{x}_p^{k+1}$, the equation
\eqref{eq:variation-averaged-strain-energy-density} can be simplified
as follows
\begin{equation}
  \label{eq:variation-averaged-strain-energy-density-simpli}
  \delta W_{\epsilon,p} =
  \frac{\delta \tens{\epsilon}_p}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I}. 
\end{equation}
Consequently we can define an equivalent critical stress at the
material point $\vec{x}_p^{k+1}$:
\begin{equation}
  \label{eq:equivalent-critical-stress}
  \tens{\sigma}_{\epsilon,p} =
  \frac{1}{m_p} \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I}, 
\end{equation}
where $m_p$ can be computed as:
\begin{equation}
  \label{eq:averaged-mass}
  m_p = \sum_{x_q^{k+1} \in B_{\epsilon}(x_p^{k+1})} m_q.
\end{equation}
This definition of the equivalent critical stress allow us to
link it with the averaged strain energy as $\delta W_{\epsilon,p} =
 \tens{\sigma}_{\epsilon,p}\ \delta\varepsilon_p$. The fracture
criterion is defined as once $\tens{\sigma}_{\epsilon,p}^{k+1}$
surpasses the tensile strength, $f_t$, the softening behaviour is
activated, which in turn causes a reduction of the initial forces as,
 \begin{equation}
   \label{eq:f-int-damaged}
   f^{int}_I = \sum_p (1 - \chi_p)\ \tens{\sigma}_{p}^{k+1} \cdot \Grad{N_{Ip}}
 \end{equation}
where $\chi_p$ is the damage variable of each material point $p$,
ranges between zero (an intact material) and one (completely failed
material points). For the case of a linear softening such the sketched
in the Figure \ref{fig:Damage-ft-wc}, it is calculated as,
 \begin{equation}
   \label{eq:damaged-variable-chi}
   1 - \chi = \frac{f_n}{f_t} = 1 - \frac{w_n}{w_c}\ \rightarrow\ \chi
   = \frac{w_n}{w_c}.
 \end{equation}

Following Bazant~\cite{Bazant83} crack-model, \cite{Navas_2017_ES}
introduced a band width $h_{\epsilon}$. The effective fracture strain
$\varepsilon_{\epsilon,f}$ is defined as the difference between the
strain at crack initialization, $\varepsilon_1(\vec{x}_p^{0})$, and
the current strain, $\varepsilon_1(\vec{x}_p^{k+1})$, for material
point $p$. Also, $\varepsilon_{\epsilon,f}$ can be represented as the
current crack opening $w_n$ within the band width,
$h_{\epsilon}$. Therefore,
\begin{equation}
  \label{eq:effective-fracture-strain}
  \varepsilon_{\epsilon,f} = \varepsilon_1(\vec{x}_p^{k+1}) -
  \varepsilon_1(\vec{x}_p^{0}) = \frac{w_n}{h_{\epsilon}}
\end{equation}
Introducing \eqref{eq:effective-fracture-strain} in
\eqref{eq:damaged-variable-chi}, the damage variable can be computed
as,
\begin{equation}
  \label{eq:damage-variable-chi-II}
\chi = \frac{\varepsilon_{\epsilon,f}\ h^{\epsilon}}{w_c}  
\end{equation}
for the case of a linear softening behaviour. In a more general case,
the damage variable can be expressed in terms of the following
variables,
\begin{equation}
  \label{eq:damage-variable-chi-III}
  \chi = \chi(\varepsilon_{\epsilon,f}, h^{\epsilon}, f_t, w_c, G_f)
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cases of study and discussion}
\label{sec:3}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{Figures/Brazilian}
  \caption{Geometry and boundary condition of the Brazilian test.}
  \label{fig:geometry-brazilian-test}
\end{figure}

@@

\begin{algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit Predictor-Corrector scheme}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Update mass matrix}:
    \begin{equation*}
      \tens{m}_{I} = N_{Ip}^{k}\ m_p,
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Predictor}:\\
    \begin{equation*}
      \vec{v}_I^{pred} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{m_I}
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions}:\\
    At the fixed boundary, set $\vec{v}_{I}^{pred} = 0$. 
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{equation*}
      \Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\
        \dot{\tens{\varepsilon}_{p}}^{k+1} = \Delta \left[ \vec{v}_{I}^{pred} \otimes
        \Grad{N_{Ip}^{k+1}} \right]^s
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Compute stress field and update damage parameter}:\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    \vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector}:\\
    \begin{equation*}
      \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\ \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \STATE \textbf{Update particles lagrangian quantities}:
    \begin{align*}
      &\vec{a}_p^{k+1} = \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Reset nodal values}
  \end{algorithmic}
\end{algorithm}

@@

\end{document}
\message{ !name(2020_EFM_MPM_Eigensoftening.tex) !offset(-630) }
